---
title: "Functional data analysis of eye pupil dilation"
author: "Paolo Girardi, Sara Scrimin and others"
date: "06/12/2019"
output: pdf_document
bibliography: C:/Users/utente/Desktop/Eye_tracker/bibliography.bib
link-citations: yes
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
\newcommand{\normb}[1]{\bigl\|{#1}\bigr\|}
\newcommand{\Ind}[1]{\mbox{\larger\textbf{1}}_{#1}}
\newcommand{\tr}{{}^{\intercal}}
\newcommand{\balpha}{\boldsymbol{\alpha}}
\newcommand{\bbeta}{\boldsymbol{\beta}}
\newcommand{\btheta}{\boldsymbol{\theta}}
\newcommand{\bgamma}{\boldsymbol{\gamma}}
\newcommand{\bdelta}{\boldsymbol{\delta}}
\newcommand{\bpsi}{\boldsymbol{\psi}}
\newcommand{\bTheta}{\boldsymbol{\Theta}}
\newcommand{\bB}{\mathbf{B}}
\newcommand{\bX}{\mathbf{X}}
\newcommand{\bY}{\mathbf{Y}}
\newcommand{\bZ}{\mathbf{Z}}
\newcommand{\ba}{\mathbf{a}}
\newcommand{\bc}{\mathbf{c}}
\newcommand{\bu}{\mathbf{u}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bw}{\mathbf{w}}
\newcommand{\bx}{\mathbf{x}}
\newcommand{\by}{\mathbf{y}}
\newcommand{\bz}{\mathbf{z}}

\abstract
In general, methods for correlating pupillary response to the cognitive activity of a subject undergoing an evaluation of cognitive activity are based on average values or analysis of peak activity.

However, in this type of analysis, we lose the temporal form of the data which is a valuable feature because of temporal scale (high temporal scale 250 Hz) or specific trend pattern. Eye tracking data contains several features:
\begin{itemize}
\item Eye point of gaze;
\item Fixation (and relative Area Of Interest -AOI-) and Saccade time periods.
\item \textbf{Eye pupil dilation}
\end{itemize}

Differently by the classical approach based on differences on ``averaged'' values over a certain period, we propose a functional data clustering approach that takes into account the entire behaviour of a eye dilation time series.


\section{1 - Functional data clustering... a gentle introduction}
The statistical problem is to cluster subjects with common temporal behaviour regarding their eye dilation.
Let $\bY_i = (Y_i (t_{i,1}),\ldots,Y_i (t_{i,m_i}))'$ a time series of $m_i$ values
collected for the subject $s_i$, $i=1,\ldots,n$. The value $m_i$ corresponds to the number of observations included in the $i-th$ time series. We suppose to observe $n$ time series and we want to cluster them in $C$ clusters.
\subsection{1.1 -  Model-based Functional data clustering}
In a mixture-model based approach to clustering the cluster membership of the i-th time seriesis represented by a latent random vector $\bZ_i=(Z_{i,1},\ldots,Z_{i,C})'$, where $Z_{i,c}=1$ if the time series belongs to the cluster $c$, $Z_{i,c}=0$ otherwise. Then a model for the complete data $(\bY_i,\bZ_i)$, $i=1,\ldots,n$ is specified in a hierarchical way.
Suppose that the $i$-th time series belongs to the cluster $c$, i.e. $Z_{i,c}=1$ and $Z_{i,c'}=0$ for c'$\ne$ c.

Given the cluster membership $\bZ_i$, we suppose that $\bY_i$ is a discrete and noised measurement of a smooth time-varying curve $f_{c}(t)$, namely
$$
Y_i(t_{i,j})=f_c(t_{i,j})+\varepsilon_{i}(t_{i,j}),\qquad j=1,\ldots,m_i.
$$
The smooth function $f_c(t)$ is described by a linear combination of $K$ B-spline basis functions $B_k(t)$, $k=1,\ldots,K$, evaluated at $K-1$ equally spaced in knots, namely
$$
f_c(t)=\sum_{k=1}^K \psi_{c,k} B_k(t)
=\gamma^\top B(t)
$$
where $\bpsi_{c}=(\psi_{c,1},\ldots, \psi_{c,K})'$ is a vector of unknown parameters.

 In view of our application, it does not seem too restrictive to assume that $\varepsilon_i(t_{i,j})$ are temporally independent Gaussian random errors with zero mean and cluster specific variance $\sigma_c^2$. It turns out that time series $\bY_1,\ldots,\bY_n$ are conditionally independent on $\bZ_1,\ldots,\bZ_n$ and 
\begin{equation}\label{eq:mod}
Y_i(t_{i,j})|\bZ_{i} \sim \mathcal{N}\left(\sum_{k=1}^K \psi_{c,k} B_k(t_{i,j}),\sigma^2_c\right), \quad j=1\ldots,m_i.
 \end{equation}
The clustering involves the membership $Z$ by means of iterative algorithm in order to get the more classification that maximized a given likelihood.

\subsection{1.2 - Two steps Functional data clustering}
A second different approach is called "2 steps - clustering". The two steps are:
\begin{itemize}
\item estimation (functional data and dimensional reduction);
\item classification.
\end{itemize}
In this attempt, we performed an unsupervised classification based on measures of the distance between the 2 curves.

We calculated a distance matrix, based on the L2 norm between two estimated curves $i$ and $j$ $f_i(t)=\sum_{k=1}^K \hat{\psi}_{i,k} B_k(t)$ and $f_j(t)=\sum_{k=1}^K \hat{\psi}_{j,k} B_k(t)$ ($\hat{\psi}$ are the estimated coefficients) is calculated as following:
\begin{equation}\label{eq:2stepsfun}
d_{i,j}= \sqrt{(\sum_{k=1}^K \hat{\psi}_{i,k} B_k(t)-\sum_{k=1}^K \hat{\psi}_{j,k} B_k(t))^2 }\approx \sqrt{\sum_{k=1}^K (\hat{\psi}_{i,k}-\hat{\psi}_{j,k})^2}  \mbox{   } \forall i \neq j,
 \end{equation}
where $d_{i_j}$ is an element of distance matrix $D$ based on the euclidean distance.

We applied to the distance matrix an algorithm in order to cluster toghether similar curves.

A summary of the statistical methods proposed for functional data clustering is reported in a review of [@jacques2014]; in the following figure the presented summary.

\includegraphics{fig2_preda.jpg}
\newpage 

\section{2 - Eye tracking dilation dataset}
We consider a dataset formed by pupil eye dilation records of 25 female childs involved in two trials consisting in the visualization of 3 images for a total of 18 seconds differing between them only for the ending scene:
\begin{itemize}
\item Trial 1: Ambiguous situation - Teacher with adverse reaction - \textbf{Child with adverse reaction};
\item Trial 2: Ambiguous situation - Teacher with adverse reaction - \textbf{Child with sadden reaction}.
\end{itemize}
\noindent\makebox[\linewidth]{\rule{\paperwidth}{0.4pt}}
\begin{table}[ht]
\begin{tabular}{c|c|c}
& \textbf{Duration of the visualizzation}&\\
Image 1: 5 secs & Image 1: 3 secs & Image 1: 10 secs\\
& \textbf{Trial 1}&\\
\hline
\includegraphics[width=0.33\textwidth]{ambigua.jpg} & \includegraphics[width=0.33\textwidth]{maestra cattiva.jpg}& \includegraphics[width=0.33\textwidth]{maestra-ostile-bambino-ostile.jpg}\\
&\textbf{Trial 2}&\\
\includegraphics[width=0.33\textwidth]{ambigua.jpg} &\includegraphics[width=0.33\textwidth]{maestra cattiva.jpg} & \includegraphics[width=0.33\textwidth]{maestra-ostile-bambino-buono.jpg}\\
\end{tabular}
\end{table}
\noindent\makebox[\linewidth]{\rule{\paperwidth}{0.4pt}}

```{r carico intro}
rm(list=ls())
setwd("C:/Users/utente/Desktop/Eye_tracker/")
### data are preprocessed (time series with 0 mean and unit standard deviation)
db_tf1<-read.csv("trial1_f.csv")
db_tf2<-read.csv("trial2_f.csv")
#imported variables
names(db_tf1)
# p_sx and d_sx are Right and Left eye dilations
# time is the temporal scale
# id is the subject ID
table(db_tf1$id)
###more than 1000 time points for each time series (18/1082= 0.017s time scale)
### but different time points 
length(table(db_tf1$time))
## Marginal distribution ofRight and Left eye dilations
boxplot(db_tf1$p_dx,db_tf1$p_sx,names=c("Right","Left"),ylab="mm")
## Right and Left eye dilations are highly correlated
cor.test(db_tf1$p_dx,db_tf1$p_sx)
plot(db_tf1$p_dx,db_tf1$p_sx,xlab="Right pupil",ylab="Left pupil")
abline(a=0,b=1,col="red",lty=2)
### create the average dilation between two eye
db_tf1$p<-apply(cbind(db_tf1$p_dx,db_tf1$p_sx),1,mean,na.rm=T)
db_tf2$p<-apply(cbind(db_tf2$p_dx,db_tf2$p_sx),1,mean,na.rm=T)




#some R function to manage plots and data
source("C:/Users/utente/Desktop/Eye_tracker/functions.R")

```
\section{Moving Average Filter}
An example of a time series on pupil dilatation is reported in the following figure.
```{r moving example}
plot(db_tf1$time[db_tf1$id=="F1301" & db_tf1$test=="t1" & db_tf1$type=="Fixation"],db_tf1$p[db_tf1$id=="F1301" & db_tf1$test=="t1"& db_tf1$type=="Fixation"],ylab="Pupil dilation",xlab="time",col=as.numeric(db_tf1$type[db_tf1$id=="F1301" & db_tf1$test=="t1"]),pch=16,ylim=c(2.4,3.1))
legend("topleft",c("Fixation","Saccade","Unclassified"),col=1:3,pch=16)
```
There is a certain grade of variability. The presence of saccade and unclassified moments lead to a very noised time series.
To overcome this and to get standardized time serie, we need to resample the time series.
We apply a Moving Average Filter in order to get the same time points for each time series as follow:
$$Yfiltered_i(t_{i,j})=\frac{\sum_{i \in W_z} Y_i(t_{i,j}) }{ \# W_z} $$
where $W_z$ is a time windows and $\bigcup_{z=1}^{\infty} W_{i}$ is the entire time window (0-18 seconds).
In addition, we scale each time series (mean=0 and var=1) because we are interested to compare only their temporal behaviour.
```{r moving average}
#set dropping first 2 seconds and endings 5 seconds
min<-2000
max<-13000
#sampling each time series at 100 ms
step<- round((max-min)/100,0)
# (data vector, subject vector, time vector, time scale, window=(min,max),scale=T)
data_sr1<-sampling_regular(db_tf1$p,db_tf1$id,db_tf1$time,step,window=c(min,max),scale=TRUE)
data_sr2<-sampling_regular(db_tf2$p,db_tf2$id,db_tf2$time,step,window=c(min,max),scale=TRUE)
# in output a list formed by a matrix of observation and a regularized time vector

# MA resampled data for the first child
plot(db_tf1$time[db_tf1$id=="F1301" & db_tf1$test=="t1" & (db_tf1$time> min & db_tf1$time< max)],scale(db_tf1$p[db_tf1$id=="F1301" & db_tf1$test=="t1" & (db_tf1$time> min & db_tf1$time< max)],scale=TRUE),type="l",ylab="Pupil dilation",xlab="time")
abline(v=c(5000,8000))
points(data_sr1$t,data_sr1$x_mat[1,],pch=16,col="red")
points(data_sr1$t,data_sr1$x_mat[1,],col="red",type="l")
legend("topleft",c("raw data","MA resampled data"),col=c(1,2),pch=c(1,2),lty=c(1,1))

#our time series 
matplot(data_sr1$t,t(data_sr1$x_mat),type="l",main="Trial 1",ylab="Pupil dilation",xlab="time")
matplot(data_sr2$t,t(data_sr2$x_mat),type="l",main="Trial 2",ylab="Pupil dilation",xlab="time")

```
Lines are overlapping. It is impossibile to observed different time patterns.
It seems hard to discover similar curves.
Next step: Smoothing!

We defined k=25 equally distributed knots for the basis of splines. A number 25 knots seems sufficient (the strategy is to have a great number of coefficients penalizing them with L2norm of applied to the second derivatives).
\begin{equation}\label{eq:smoothing pen}
\widehat{\psi_{c,k}}=argmin \sum_{t=1}^T \left( Y_it-\sum_{k=1}^K \psi_{c,k} B_k(t)\right)^2+\lambda \int  \frac{\partial^2 \psi_{c,k} B_k(t)}{\partial t^2}
 \end{equation}
where $\lambda$ is the smoothing parameter that penalized roughness curves.
The number of smoothing $\lambda$ is chosen by means of the GCV (Generalized Cross validation) criterion. 

```{r smooth }
# we adopt a smoothing strategy, high number of knots (15) 
# penalizing with L2 norm on second derivatives of coefficients
# GCV criterion to select the best lambda 

l<-exp(seq(10,15,0.25))

gcv1<-NULL
for(i in 1:length(l)){
gcv1<-c(gcv1,fun_register(data_sr1,lambda=l[i], knots=25,plot=FALSE,register=FALSE)$gcv)
}
graphics::plot(gcv1,ylab="GCV",xlab="lambda value")
which.min(gcv1)

gcv2<-NULL
for(i in 1:length(l)){
gcv2<-c(gcv2,fun_register(data_sr2,lambda=l[i], knots=25,plot=FALSE,register=FALSE)$gcv)
}
graphics::plot(gcv2,ylab="GCV",xlab="lambda value")
which.min(gcv2)


# coefficient estimation 
data_reg1<-fun_register(data_sr1,lambda=l[which.min(gcv1)], knots=25,plot=FALSE,register=FALSE)$data_reg
data_reg2<-fun_register(data_sr2,lambda=l[which.min(gcv2)], knots=25,plot=FALSE,register=FALSE)$data_reg

#example of estimated curve
plot(data_sr1$t,data_sr1$x_mat[1,],ylab="Pupil dilation",xlab="time")
plot(data_reg1[1],add=T,col=2)
legend("topleft",c("Curve estimated"),col=c(2),lty=1)

```

\section{Number of clusters}
The choice of the number of clusters is a crucial point: we select the number of cluster by means of the Gap Statistics [@tibshirani2001estimating], using PAM algorithm applying the euclidean distance (L2-norm) to the B-splines coefficients (the L2 norm applied to the curves or to the Bspline coefficient is the same). 
```{r number of clusters}
library(cluster)
#euclidean distance matrix
dist_mat1<-dist(t(data_reg1$coefs))
dist_mat2<-dist(t(data_reg2$coefs))

pam1 <- function(x,k) list(cluster = pam(x,k, cluster.only=TRUE,diss=T))
mod_sel<-clusGap(as.matrix(dist_mat1), FUN = pam1, K.max = 8, B =500,spaceH0 = "original")
#useTibs2001SEmax method and SE.factor=1
print(mod_sel, method="Tibs2001SEmax",SE.factor=1)
plot(mod_sel)


mod_sel<-clusGap(as.matrix(dist_mat2), FUN = pam1, K.max = 8, B=500,spaceH0 = "original")
#uso FirstSEmax method
print(mod_sel, method="Tibs2001SEmax",SE.factor=1)
plot(mod_sel)

```




For both trials, Gap statistic suggests 3 clusters. At the end we cluster time series and make appropriate graphs.

```{r clustering}


#two clusterings
clust1<-pam(as.matrix(dist_mat1),3,diss=T)$clust
clust2<-pam(as.matrix(dist_mat2),3,diss=T)$clust
table(clust1)
table(clust2)
table(clust1,clust2)
```
\subsection{Results}
In the first trials we classified 16 time series in the first group, 6 in the second one and 3 in the last group
In the second trials we classified 10, 8 and 7 curves in the group 1, 2 and 3 respectively.
We can comment in function of the different behaviours on temporal profile in each group by trial membership.
```{r fitted curves}

plot_cluster_curves2(data_reg1,clust1)
legend("topleft",c("cluster 1","cluster 2","cluster 3"),lty=1, col=1:3)
abline(v=c(5000,8000))

plot_cluster_curves2(data_reg2,clust2)
legend("topleft",c("cluster 1","cluster 2","cluster 3"),lty=1, col=1:3)
abline(v=c(5000,8000))



par(mfrow=c(3,2),mar = c(2, 2, 2, 2),mgp=c(5,1,0))
for (i in 1:3){ 
plot(data_reg1[clust1==i],lty=1)
title(paste("trial 1 - cluster",i, sep=""))
plot(data_reg2[clust2==i],lty=1)
title(paste("trial 2 - cluster",i, sep=""))
}
par(mfrow=c(1,1))



for (i in 1:max(clust1)){ 
plot_cluster_curves2(data_reg1[i %in% clust1],clust1==i,lwd=2)
abline(v=c(5000,8000))
title(paste("trial 1 cluster",i))
for(j in 1:table(clust1)[i]) points(data_sr1$t,data_sr1$x_mat[clust1==i,][j,],cex=0.5)
}

for (i in 1:max(clust2)){ 
plot_cluster_curves2(data_reg2[i %in% clust2],clust2==i,lwd=2) 
abline(v=c(5000,8000))
title(paste("trial 2 cluster",i))
for(j in 1:table(clust2)[i]) points(data_sr2$t,data_sr2$x_mat[clust2==i,][j,],cex=0.5)
}

```
\section{Conclusions}
To be continued....


\section{Other}

R-packages for EyeTrackerData

http://www.eyetracking-r.com/

#References
